---
title: "YOUR LESSON TITLE HERE". Example: "Data Wrangling: Cleaning and Tidying Raw Data"
subtitle: "YOUR LESSON SUBTITLE HERE". Example: "Transforming and Refining Raw Sports Data for Insightful Analysis"
abstract: "YOUR LESSON ABSTRACT HERE". Example: "Dive into the intricacies of data wrangling to unveil the true potential of sports data. This lesson is dedicated to transforming raw datasets into a structured, analysis-ready format, essential for extracting actionable insights. We navigate through data cleaning and tidying, employing tools from R's `tidyverse` to refine and reshape datasets for sports and exercise science applications. Designed for sports analysts and practitioners, this lesson offers a hands-on approach to mastering the art of data wrangling, focusing on practical techniques for dealing with common data challenges. From parsing strings to reshaping data formats, you'll emerge equipped with the skills to turn raw numbers into compelling stories and evidence-based decisions."
format:
  html:
    code-fold: false
    toc: true
    toc_depth: 4
    toc-title: Contents
    smooth-scroll: true
    number-sections: true
    css: styles.css
  pdf: default
execute:
  warning: false
editor: visual
editor_options: 
  chunk_output_type: inline
---

<!-- Lesson's keywords here. Example below -->
::: {.callout-tip title="Keywords"}
Data wrangling, data cleaning, data tidying, data transformation, data organization, data manipulation, tibble, data analysis, sports analytics, R, tidyverse, dplyr, tidyr, stringr, visdat, speedsR.
:::

<!-- Specify the lesson's difficulty level here. -->
::: {.callout-note title="Lesson's Level"}
The level of this lesson is categorized as [BRONZE]{style="color:#CD7F32"}.
:::

<!-- Identify the lesson's main idea here. Example below. -->
::: {.callout-tip title="**Lesson's Main Idea**"}
-   Data wrangling, cleaning, and tidying form the bedrock of meaningful sports analytics, transforming raw datasets into structured, insightful sources.
-   Mastering these processes in R's `tidyverse` empowers practitioners to reveal hidden patterns and insights essential for data-driven decision-making in sport and exercise science science.
:::

<!-- Sometimes you might need to provide a disclaimer for the lesson like the one below. If there's no disclaimer in your lesson, delete this.  -->
::: {#disclaimer .callout-warning title="Dataset Featured In This Lesson"}
**Disclaimer**: The dataset featured in this lesson is entirely fictitious and has been artificially created for pedagogical purposes. Any resemblance to actual persons, living or deceased, or real-world data is purely coincidental and not intended.
:::

<!-- Outline what a learner might expect to get out of the lesson here. Example below. -->
# Learning Outcomes

By the end of this lesson, you will have honed the ability to:

-   **Manage and Concatenate Data Files**: Skillfully manage and concatenate multi-file datasets, harnessing the power of R to streamline data preparation processes in sport analytics.

-   **Execute Data Wrangling Techniques**: Apply robust data wrangling methods to transform raw datasets into analyzable formats, enabling deeper insights in sport and exercise science.

-   **Organize, Reshape, and Order Data Effectively**: Implement data tidying techniques to structure and reshape datasets, as well as order temporal data, facilitating easier manipulation, modeling, and visualization.

-   **Ensure Data Precision**: Master data cleaning strategies to rectify inaccuracies and standardize datasets, ensuring precision and consistency for accurate sports data analysis.

# Introduction: Navigating Data Refinement in Sport and Exercise Science

<!-- Introduction subsections here -->


<!-- Might be useful to outline main tools used for the lesson's subject. Example below. -->
# Tools for Data Wrangling, Cleaning, and Tidying

## `dplyr` package in R

`dplyr` is a fundamental package in R for data wrangling, cleaning, and tidying. In our lesson, we tap into a small yet powerful portion of `dplyr`'s capabilities. We use functions like `distinct()` to remove duplicates and `gsub()` for fine-tuning string values, which are key for tidying column names. Additionally, `rename()` plays a crucial role in making our data more readable and consistent. These functions are just examples; `dplyr` offers a broad range of tools that are incredibly useful for cleaning and organizing data, especially vital for sport science data analysis.

## `tidyr` package in R

The `tidyr` package is a key player in the R environment for data tidying. It offers specialized functions like `gather()` for reshaping data from wide to long formats, a critical step for data analysis. We also utilize `tidyr`'s `separate()` function, which allows us to split complex column entries into simpler, more manageable parts. Used alongside `str_sub()` from the `stringr` package, `tidyr` boosts our ability to efficiently clean and structure data. This combination of tools makes `tidyr` an essential package for data scientists, especially in sports science where data clarity and precision are paramount.

Both `dplyr` and `tidyr`, key components of the `tidyverse` collection of R packages designed for data science, play a crucial role in transforming raw, messy, and cluttered datasets into clean, structured formats ready for analysis. Their combined functionalities tackle a wide range of data wrangling tasks, establishing them as essential tools in the data science toolkit, especially for those in sports science and related fields.


<!-- Main body of the lesson -->
# Wrangling, Cleaning, and Tidying an 'Unclean' Dataset

<!-- It's important to explicitly equip a learner with all the libraries and installations they might need for this lesson. -->
## Install and Load Libraries

Most of the libraries required for this lesson are readily available on CRAN and can be installed using the command `install.packages("NameOfPackage")`, if they are not already installed on your computer:

```{r, eval=FALSE}
install.packages(c("dplyr", "tidyr", "readr", "stringr", "visdat"), repos="https://cloud.r-project.org/")
```

Another library required for this lesson, `speedsR`, is specifically designed for the AIS SPEEDS project. This package, not available on CRAN, offers an extensive collection of sports-specific datasets, including the one used in this lesson. `speedsR` can be downloaded from the AIS SPEEDS GitHub repository and integrated into any R-based Integrated Development Environment (IDE). A detailed instruction on how to install and load `speedsR` is provided in the [dedicated section of our website](https://speeds.quarto.pub/speeds/speedsR_data_package.html).

After you have installed all the required packages, you can load them into your current R session as follows:

```{r}
#| message: false
#| results: false

rm(list = ls()) # clear the workspace

library(dplyr) # provides functions for manipulating datasets
library(tidyr) # tools for tidying data, reshaping and restructuring datasets
library(readr) # offers fast and user-friendly functions to read tabular data into R
library(stringr) # simplifies the process of working with strings, providing consistent and easy-to-understand functions
library(visdat) # enables visual inspection of data quality and structure
library(speedsR) # collection of sports-specific benchmark datasets - part of the AIS SPEEDS project
```

## Load Datasets From The `speedsR` Package 

<!-- Main body of the lesson continues -->


<!-- Challenge the learner here - suggest some exercises following the lesson's content. -->
::: {.callout-tip title="Challenge for the Learner"}
Apply your data wrangling skills to refine and analyze a new dataset:

1.  Identify a sports dataset of interest and perform initial data cleaning steps, such as standardizing column names and handling missing values.
2.  Utilize string parsing to extract useful information from complex data entries, and reshape the dataset from wide to long format.
3.  Craft a visualization of the cleaned dataset to reveal an interesting trend or insight, using the principles of tidy data you've learned.
:::

<!-- Summary for the lesson. Example below. -->
# Conclusion and Reflection

In this lesson, we've honed the essential skills of data wrangling, cleaning, and tidying, crucial for insightful sports data analytics. We've transformed raw datasets into structured, analysis-ready formats, crucial for unearthing patterns and driving data-informed decisions.

The journey through R's `tidyverse` has not only shown us how to effectively clean and restructure data but also underscored the importance of precise data transformation for reliable analysis. Applying these data refinement techniques to the diverse datasets encountered in sport and exercise science empowers you to unlock analytical potential. Each step in mastering these skills paves the way to new insights and advancements.


<!-- Offer learners an opportunity to test their knowledge with a little quiz here. Example below. -->
# Knowledge Spot-Check

::: {.callout-tip title="**What is the primary goal of data wrangling in sports data analytics?** <br> <br> A) To create visually appealing reports. <br> B) To ensure data is formatted for easy entry into spreadsheets. <br> C) To transform and clean data for accurate analysis. <br> D) To learn programming in R and Python. <br> <br> Expand to see the correct answer." collapse="true"}
The correct answer is C) To transform and clean data for accurate analysis.
:::

::: {.callout-tip title="**Which R package provides functions for both data wrangling and tidying?** <br> <br> A) ggplot2 <br> B) shiny <br> C) readr <br> D) tidyverse <br> <br> Expand to see the correct answer." collapse="true"}
The correct answer is D) tidyverse.
:::

::: {.callout-tip title="**When preparing data for analysis, why is it important to standardize column names?** <br> <br> A) To enable merging with other datasets. <br> B) To prevent errors due to case sensitivity. <br> C) To facilitate sorting of columns alphabetically. <br> D) To make the dataset visually appealing. <br> <br> Expand to see the correct answer." collapse="true"}
The correct answer is B) To prevent errors due to case sensitivity.
:::

::: {.callout-tip title="**What does the R function 'gsub()' do in the context of data cleaning?** <br> <br> A) Summarizes data statistics. <br> B) Generates subplots for data visualization. <br> C) Replaces patterns in strings. <br> D) Groups similar data points. <br> <br> Expand to see the correct answer." collapse="true"}
The correct answer is C) Replaces patterns in strings.
:::

::: {.callout-tip title="**Why is the 'long' format often preferred over the 'wide' format in data analysis?** <br> <br> A) It is more compact and requires less disk space. <br> B) It typically allows for more efficient computation. <br> C) It facilitates easier filtering, summarizing, and visualization. <br> D) It is the only format supported by modern analytics tools. <br> <br> Expand to see the correct answer." collapse="true"}
The correct answer is C) It facilitates easier filtering, summarizing, and visualization.
:::
