---
title: "Tutorial Template R"
description: "Exploring the Iris Dataset"
format:
  html:
    code-fold: false
---

## Introduction

This is a template tutorial. The Iris dataset is explored here as an example in R. 

## Loading Iris Dataset

In R, the Iris dataset comes preloaded, so we can directly start working with it.

## Previewing the Data

Let's take a look at the first few rows of our dataset.

```{r}
head(iris)
```

## Descriptive Statistics

We'll obtain some descriptive statistics to understand the dataset better.

```{r}
summary(iris)
```

## Checking for Missing Values

Let's make sure that our dataset doesn't contain any missing values.

```{r}
sum(is.na(iris))
```

## Checking the Dimensions of the Dataset

We'll print the dimensions of our dataset to understand its structure.

```{r}
print(paste("Dimensions of the dataset:", paste(dim(iris), collapse = " x ")))
```

## Visualizing Pairwise Relationships

Finally, we'll create a pair plot to visualize pairwise relationships between variables, separated by the species.

```{r}
library(ggplot2)
pairs(iris[1:4], main = "Iris Data", pch = 21, bg = c("red", "green3", "blue")[unclass(iris$Species)])
```

## Conclusion

Through this tutorial, we explored the Iris dataset, obtained descriptive statistics, and visualized pairwise relationships. This is a foundational step in any data analysis project.